Bayesian regression facilitates quantitative modelling of cell metabolism

### 2023-11-01 paper/main.qmd
Paper comments with Lars

---
title: "Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism, 
          multiomics integration, ordinary differential equations,
          regulatory anlaysis
abstract: >
  This paper presents Maud, a command-line application that implements Bayesian
  statistical inference for kinetic models of biochemical metabolic reaction
  networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Our paper
  reviews the existing options in this area, presents a case study illustrating
  how Maud can be used to analyse a metabolic network and explains the
  biological, statistical and computational design decisions underpinning Maud.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Nicholas Luke Cowie"
    affiliations: 
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
    attributes:
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Lars Keld Nielsen"
    affiliations: 
      - name: The University of Queensland
        department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
        city: St Lucia
        country: Australia
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

Our application Maud provides the first Bayesian kinetic model to combine
biologically realistic mechanistic accuracy---including accurate rate laws,
post- translational modification and thermodynamics---with fast, robust
MCMC sampling using adaptive Hamiltonian Monte Carlo. Further, Maud is a
general-purpose application that can be used to fit a wide range of Bayesian
kinetic models.

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared to the
corresponding marginal prior distributions, even though these parameters’
joint posterior distribution contained enough information to make accurate out
of sample predictions. In some cases, there were two-dimensional correlations
such as the one shown in the bottom left of @fig-posterior; in this case the
marginal distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to reveal the
underlying correlation structure, as seen in the bottom-right plot in figure
@fig-posterior, which shows two marginally independent parameters. This kind of
marginal independence does not mean that the parameters are independent in the
full joint posterior distribution.

@@

Marginal posterior distributions from our main model run. (**Top left**)
Comparison of posterior predictive intervals with simulated flux measurement
values. All the flux measurements are within the predictive intervals,
indicating a good fit. (**Top right**) Comparison of posterior predictive
intervals with simulated concentration measurement values. These also show
a good fit. (**Bottom left**) Pairwise marginal posterior distribution
for two pairwise correlated parameters, namely $K_m^{AHC1,hcys-L}$ and
$K_m^{AHC1_ahcys}$. (**Bottom right**) Pairwise marginal posterior distribution
for two pairwise-uncorrelated parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table 1 in
this paper's supporting information for a description of all these parameters
and their dimensions. Note that some quantities in Maud’s kinetic model are
not treated as parameters: for example, temperatures, compartment volumes and
the formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values can be
used. Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest of
simplicity.

>>>>

---
title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism, 
          multiomics integration, ordinary differential equations,
          regulatory anlaysis
abstract: >
  This paper presents Maud, a command-line application that implements Bayesian
  statistical inference for kinetic models of biochemical metabolic reaction
  networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Our paper
  reviews the existing options in this area, presents a case study illustrating
  how Maud can be used to analyse a metabolic network and explains the
  biological, statistical and computational design decisions underpinning Maud.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
      - id: biosustain
        name: DTU
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Nicholas Luke Cowie"
    orcid: 0000-0003-3367-2105
    affiliations: 
      - ref: biosustain
    attributes:
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Lars Keld Nielsen"
    orcid: 0000-0001-8191-3511
    affiliations: 
      - id: uq
        name: The University of Queensland
        department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
        city: St Lucia
        country: Australia
      - ref: biosustain
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

Our application Maud provides a Bayesian kinetic model that combines
biologically realistic mechanistic accuracy---including accurate rate laws,
post- translational modification and thermodynamics---with fast, robust
MCMC sampling using adaptive Hamiltonian Monte Carlo. Further, Maud is a
general-purpose application that can be used to fit a wide range of Bayesian
kinetic models.

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared to the
corresponding marginal prior distributions, even though these parameters’
joint posterior distribution contained enough information to make accurate out
of sample predictions. In some cases, there were two-dimensional correlations
such as the one shown in the bottom left of @fig-posterior; in this case the
marginal distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to reveal the
underlying correlation structure, as seen in the bottom-right plot in figure
@fig-posterior, which shows two marginally independent parameters.

@@

Marginal posterior distributions from our main model run. (**Top left**)
Comparison of posterior predictive intervals with simulated flux measurement
values. All the flux measurements are within the predictive intervals,
indicating a good fit. (**Top right**) Comparison of posterior predictive
intervals with simulated concentration measurement values. These also show
a good fit. (**Bottom left**) Pairwise marginal posterior distribution
for two correlated parameters, namely $K_m^{AHC1,hcys-L}$ and
$K_m^{AHC1_ahcys}$. (**Bottom right**) Pairwise marginal posterior distribution
for two uncorrelated parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table 1 in
this paper's supporting information for a description of all these parameters
and their dimensions.

### 2023-11-01 paper/main.tex
Paper comments with Lars

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  pdfkeywords={Bayesian inference, kinetic models of cell metabolism,
multiomics integration, ordinary differential equations, regulatory
anlaysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\author{Teddy Groves}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}

@@

\email{tedgro@biosustain.dtu.dk}
\author{Nicholas Luke Cowie}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}

@@

\author{Lars Keld Nielsen}
\affiliation{Australian Institute for Bioengineering and Nanotechnology
(AIBN), The University of Queensland, St Lucia, Australia}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}

@@

\title[]{Maud: Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared to the
corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
Figure~\ref{fig-posterior}; in this case the marginal distribution of
the two parameters is roughly banana shaped. More commonly, however,
two-dimensional pair plots were insufficient to reveal the underlying
correlation structure, as seen in the bottom-right plot in figure
Figure~\ref{fig-posterior}, which shows two marginally independent
parameters. This kind of marginal independence does not mean that the
parameters are independent in the full joint posterior distribution.

@@

\caption{\label{fig-posterior}Marginal posterior distributions from our
main model run. (\textbf{Top left}) Comparison of posterior predictive
intervals with simulated flux measurement values. All the flux
measurements are within the predictive intervals, indicating a good fit.
(\textbf{Top right}) Comparison of posterior predictive intervals with
simulated concentration measurement values. These also show a good fit.
(\textbf{Bottom left}) Pairwise marginal posterior distribution for two
pairwise correlated parameters, namely \(K_m^{AHC1,hcys-L}\) and
\(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
posterior distribution for two pairwise-uncorrelated parameters, namely
\(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table 1 in this paper's supporting information for a
description of all these parameters and their dimensions. Note that some
quantities in Maud's kinetic model are not treated as parameters: for
example, temperatures, compartment volumes and the formation energy of
water. Maud treats these quantities as if they were known precisely:
they can be configured by the user or default values can be used.
Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest
of simplicity.

>>>>

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  pdfkeywords={Bayesian inference, kinetic models of cell metabolism,
multiomics integration, ordinary differential equations, regulatory
anlaysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\author{Teddy Groves}
\affiliation{The Novo Nordisk Foundation Center for
Biosustainability, DTU, Kongens Lyngby, Denmark}

@@

\email{tedgro@biosustain.dtu.dk}
\author{Nicholas Luke Cowie}
\affiliation{The Novo Nordisk Foundation Center for
Biosustainability, DTU, Kongens Lyngby, Denmark}

@@

\author{Lars Keld Nielsen}
\affiliation{The Novo Nordisk Foundation Center for
Biosustainability, DTU, Kongens Lyngby, Denmark}
\affiliation{Australian Institute for Bioengineering and Nanotechnology
(AIBN), The University of Queensland, St Lucia, Australia}

@@

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared to the
corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
Figure~\ref{fig-posterior}; in this case the marginal distribution of
the two parameters is roughly banana shaped. More commonly, however,
two-dimensional pair plots were insufficient to reveal the underlying
correlation structure, as seen in the bottom-right plot in figure
Figure~\ref{fig-posterior}, which shows two marginally independent
parameters.

@@

\caption{\label{fig-posterior}Marginal posterior distributions from our
main model run. (\textbf{Top left}) Comparison of posterior predictive
intervals with simulated flux measurement values. All the flux
measurements are within the predictive intervals, indicating a good fit.
(\textbf{Top right}) Comparison of posterior predictive intervals with
simulated concentration measurement values. These also show a good fit.
(\textbf{Bottom left}) Pairwise marginal posterior distribution for two
correlated parameters, namely \(K_m^{AHC1,hcys-L}\) and
\(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
posterior distribution for two uncorrelated parameters, namely
\(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table 1 in this paper's supporting information for a
description of all these parameters and their dimensions.
